Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-01-29 doc/papers/iop2014/paper.tex
Take into account other reviews.

\authorinfo{Serge Guelton}
           {T{\'e}l{\'e}com Bretagne}
           {serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}
           {INRIA/MOAIS}
           {pierrick.brunet@inria.fr}
\authorinfo{Alan Raynaud}
           {T{\'e}l{\'e}com Bretagne}
           {alan.raynaud@telecom-bretagne.eu}
\authorinfo{Mehdi Amini}
           {SILKAN}
           {mehdi.amini@silkan.com}
\authorinfo{Adrien Merlini}
           {T{\'e}l{\'e}com Bretagne}
           {adrien.merlini@telecom-bretagne.eu}
\authorinfo{Eliott Coyac}
           {T{\'e}l{\'e}com Bretagne}
           {eliott.coyac@telecom-bretagne.eu}

@@

    In addition to these compilation steps, Pythran provides a C++ runtime 
    library that
    leverages the C++ STL to provide generic containers, and the Numeric 
    % FIXME (phrase trop longuuuuuuuuue) 
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.
    % FIXME Expression template is not a "feature" of the language IMO
    % It is more a classical "use" of the language

    % FIXME  (explain that annotations can be added to the code?)
    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, % FIXME (cut sentence) 
    without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr or numba, without any change to
    the original code. 
    % It may be a little too ... untactful?

@@

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high, % FIXME cut sentence
due to numerous dynamic lookup and interpretation cost inherent
in high level languages. Additionally, the Python compiler does not perform
optimization on the byte code, while scientific applications are first-class
candidates for many of them.

Since scientific applications are often told to spend 90\% of their time in
10\% of the code, it is natural to focus on those computation-intensive pieces 
of code. The aim % FIXME (of what) 
may not be to optimize the full Python application, but rather a subset of the application.

@@

One approach, used by Cython \cite{cython2010} is to suppress the interpretation
overhead by translating Python Programs to C programs calling the Python C
API\cite{pythoncapi}. More recently, Nuitka\cite{nuitka}  has taken the same
approach using C++ as a back-end. Going a step further Cython also provides an
hybrid C/Python language that can efficiently be translated into C code. It 
relies both on the Python C API and on plain C.
ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python programs
into C++, without any call to the Python C API, but fails to compile dynamic
codes or codes for which static type inference is impossible.

The alternate approach consists in writing a JIT compiler, embedded into the
interpreter, to dynamically turn the computation intensive parts into native
code. This is what the \texttt{numexpr} module\cite{numexpr} does for Numpy 
expressions, JIT-compiling them from a string representation to native code.
Numba\cite{numba} and Parakeet\cite{parakeet2012} extend this approach to
Numpy-centric applications while PyPy\cite{pypy2009} applies it to the whole
language - including the Numpy module through the in-progress \texttt{numpypy}
branch.

Except for PyPy, none of these compilers apply any of the
static optimization techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled language such as C or C++.
Translators to statically compiled languages do take advantage of them
indirectly, but the quality of the generated code may prevent advanced
optimizations, such as vectorization, %FIXME too long
while they are available at higher level,
i.e. at the Python level. Taking into account the specificities of the Python
language can unlock many new transformations. %FIXME sentence without subject
For instance, PyPy automatically
converts the \texttt{range} builtin into \texttt{xrange} by using a dedicated
structure called \texttt{range-list}.

@@

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is then translated to polymorphic C++ code through the
generation of templated classes in place of Python functions. These
classes are either used to generate regular C++ % FIXME (not clear what do you mean?) PB : Better?
code or instantiated for given types to generate a native Python module.

@@

    \item the closure of nested functions is statically computed to turn nested
        functions into global functions taking the closure elements as
        parameters, and the initial definition is replaced by a \emph{partial
        function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;
        % FIXME (this sentence makes my head hurt :s maybe cut into smaller sentences)

@@

    \item \texttt{try \dots finally} constructs are turned into nested \texttt{try \dots except} blocks;
        % FIXME (Is the difference between construct and block obvious for intended readers?)

@@

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its life
time, but it cannot be assigned to a string if it has already been assigned to
an integer. % FIXME (not sure if clear enough) 
As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected, which is very
similar to Python's duck typing, except that it is checked at compile time, as
% FIXME (tooooooooooooooooooooooo long) 
illustrated by the following implementation of a generic dot product in Python:

@@

Although far more verbose % FIXME  (don't think you can use verbose this way)
than the Python version, the C++ version also uses some
form of structural typing: the only assumptions these two versions make are that
\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and 
that the result of the multiplication is accumulatable.

@@

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions written in the
module. % FIXME (not only the functions called by top-level functs????)
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11\cite{isocxx11} language and leverage recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.8.2 and clang++ 3.3.

@@

When turning Python AST into Pythran IR, nested functions are turned into global
functions taking their expanded closure as parameter. This closure is computed 
using the information provided by the \texttt{Globals} analysis which 
statically computes the state of the dictionary of globals, and 
\texttt{ImportedIds} that computes the set of % FIXME (too long)
identifiers used by an instruction but not declared in this instruction. For
instance in the following snippet:

@@

Using the proper library, the \cpp language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the \cpp Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural choice to represent Python's \texttt{long} in \cpp. NT2 provides a generic
vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
the use of vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

@@

    \item Expression templates\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregate all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        and hence increase the impact of the two following optimizations.
        % what about binary func?

@@

    \item Loop parallelization through OpenMP\cite{openmp3.1}. Numpy expression
        computation do not carry any loop-dependency. They are perfect
        candidates for loop parallelization, especially after the expression
        templates aggregation, as OpenMP generally performs better on loops
        with higher computation intensity that masks the scheduling overhead.

@@

The OpenMP standard\cite{openmp3.1} is a widely used solution for Fortran, C
and C++ to describe loop-based and task-based parallelism. It consists of a few
directives attached to the code, directives that describe parallel loops and parallel code
sections in a shared memory model.

@@

The following listing gives a simple example of explicit loop-based 
parallelism. OpenMP 3.0 task-based parallelism form is also supported, the
interested reader may refer to~\cite{pyhpc2013} for more details.

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    #omp parallel for reduction(+:hits)
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}
% why #? I thought string was enough?

The loop is flagged as parallel, performing a reduction using the \texttt{+}
operator on the \texttt{hits} variable. Variables \texttt{x}, \texttt{y} and
\texttt{dist} are automatically made local by Pythran after a
\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, \ie{local to
a thread and not shared with other threads, by OpenMP}.

% FIXME (Ca semble tomber a plat, pas de resultats, pas de conclusion...)

Next section performs an in-depth comparison of Pythran with others Python
optimizers: Numba, PyPy, ShedSkin and numexpr.

@@

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expression is still note mature for numpy expression, but there is a
strong ungoing effort of the community to fix this.

@@

The ``blacksholes'' application is a benchmarks that showcases the
translation cost from the Python space to native space. Indeed we forced the
use of a \texttt{list} as input of the kernel instead of a numpy array. As a
consequence there is a costly translation step from a structure that contains
an array of pointers to generic Python objects to a dense structure involved.
Results are shown in Table~\ref{tbl:black}. They show that conversion code is
less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
it does not suffer from a similar conversion penality.

@@

This paper gives an overview of the compilation flow, the analyses involved and
the optimizations used. It also compares the performance of compiled Pythran
modules against CPython and other optimizers: ShedSkin, PyPy and numexpr.

To conclude, limiting Python to a statically typed subset does not hinders the
expressivity when it comes to scientific or mathematic computations, but makes
it possible to use a wide variety of classical optimizations to help Python
match the performances of statically compiled language. Moreover, one can use
high level information to generate efficient code that would be difficult to
write for the average programmer.

@@

This project has been partially funded by the CARP Project and the SILKAN
Company.

>>>>

\authorinfo{Serge Guelton}
           {T{\'e}l{\'e}com Bretagne}
           {serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}
           {INRIA/MOAIS}
           {pierrick.brunet@inria.fr}
\authorinfo{Mehdi Amini}
           {SILKAN}
           {mehdi.amini@silkan.com}
\authorinfo{Adrien Merlini}
           {T{\'e}l{\'e}com Bretagne}
           {adrien.merlini@telecom-bretagne.eu}
\authorinfo{Xavier Corbillon}
           {T{\'e}l{\'e}com Bretagne}
           {xavier.corbillon@telecom-bretagne.eu}
\authorinfo{Alan Raynaud}
           {T{\'e}l{\'e}com Bretagne}
           {alan.raynaud@telecom-bretagne.eu}

@@

    In addition to these compilation steps, Pythran provides a C++ runtime 
    library that
    leverages the C++ STL to provide generic containers, and the Numeric 
    % FIXME (phrase trop longuuuuuuuuue) 
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical idioms such as expression templates.

    % FIXME  (explain that annotations can be added to the code?)
    Unlike the Cython approach , Pythran input code remains compatible with the
    Python interpreter. Output code is generally as efficient as the annotated
    Cython equivalent, if not more, but without the backward compatibility
    loss.

@@

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high. % FIXME cut sentence It is partly due to the interpration cost,
but also to Python's dynamic binding. Additionally, the Python compiler does
not perform any optimization on the byte code, while scientific applications
are first-class candidates for many of them.

Since scientific applications are often told to spend 90\% of their time in
10\% of the code, it is natural to focus on those computation-intensive pieces 
of code. The goal of an optimizing compiler % FIXME (of what) 
may not be to optimize the full Python application, but rather a subset of the application.

@@

One approach, used by Cython~\cite{cython2010} is to suppress the interpretation
overhead by translating Python Programs to C programs calling the Python C
API~\cite{pythoncapi}. More recently, Nuitka~\cite{nuitka}  has taken the same
approach using C++ as a back-end. Going a step further Cython also provides an
hybrid C/Python language that can efficiently be translated into C code. It 
relies both on the Python C API and on plain C.
ShedSkin~\cite{shedskin2006} translates implicitly strongly typed Python programs
into C++, without any call to the Python C API, but fails to compile dynamic
codes or codes for which static type inference is impossible.

The alternate approach consists in writing a JIT compiler, embedded into the
interpreter, to dynamically turn the computation intensive parts into native
code. This is what the \texttt{numexpr} module~\cite{numexpr} does for Numpy 
expressions, JIT-compiling them from a string representation to native code.
Numba~\cite{numba} and Parakeet~\cite{parakeet2012} extend this approach to
Numpy-centric applications while PyPy~\cite{pypy2009} applies it to the whole
language - including the Numpy module through the in-progress \texttt{numpypy}
branch.

Except for PyPy, none of these compilers apply any of the static optimization
techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled language such as C or C++.  Translators to
statically compiled languages do take advantage of them indirectly, but the
quality of the generated code may prevent advanced optimizations. For instance
GCC hardly vectorizes the output of Cython, while the array semantic is
avaialable at the Python level. Taking into account the specificities of the
Python language can unlock many new transformations.  For instance, PyPy
automatically converts the \texttt{range} builtin into \texttt{xrange} by using
a dedicated structure called \texttt{range-list}.

@@

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is then translated to polymorphic C++ code through the
generation of templated classes in place of Python functions. These classes are
either used from regular C++ applications, or instantiated for given types to
generate a native Python module.

@@

    \item the closure of nested functions is statically computed to turn nested
        functions into global functions taking the closure elements as
        parameters. The initial definition is replaced by a \emph{partial
        function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;

@@

    \item \texttt{try \dots finally} instriuction are turned into nested \texttt{try \dots except} instructions;

@@

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its life
time, but it cannot be assigned to a string if it has already been assigned to
an integer.  As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected. This technique is
very similar to Python's duck typing, except that it is checked at compile
time, as illustrated by the following implementation of a generic dot product
in Python:

@@

Although more verbose than the Python version, the C++ version also uses some
form of structural typing: the only assumptions these two versions make are
that \texttt{L0} and \texttt{L1} are iterable, their content can be multiplied
and that the result of the multiplication is accumulatable.

@@

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions called in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

The last step involves a template library, called \texttt{pythonic} that 
contains a polymorphic implementation of many functions of the Python standard
library in the form of C++ template functions. Several optimizations, most
notably using expression template, are delegated to this library. Pythran relies
on the C++11~\cite{isocxx11} language and leverage recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic
templates.
As a consequence it requires a C++11 compatible compiler for the native code
generation. Boost.Python~\cite{boostpython2007} is used for the automatic
Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
4.8.2 and clang++ 3.3.

@@

When turning Python AST into Pythran IR, nested functions are turned into
global functions taking their expanded closure as parameter. This closure is
computed using the information provided by the \texttt{Globals} and
\texttt{ImportedIds} analyse. The former statically computes the state of the
dictionary of globals, and the latter computes the set of identifiers used by
an instruction but not declared in this instruction. For instance in the
following snippet:

@@

Using the proper library, the \cpp language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the \cpp Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural choice to represent Python's \texttt{long} in \cpp. NT2 provides a generic
vector library called \texttt{boost.simd}~\cite{esterie2012boost} that enables
the use of vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

@@

    \item Expression templates~\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregate all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        and hence increase the impact of the two following optimizations.
        % what about binary func?
        % ufunc -> micro function, in numpy ;-)

@@

    \item Loop parallelization through OpenMP~\cite{openmp3.1}. Numpy expression
        computation do not carry any loop-dependency. They are perfect
        candidates for loop parallelization, especially after the expression
        templates aggregation, as OpenMP generally performs better on loops
        with higher computation intensity that masks the scheduling overhead.

@@

The OpenMP standard~\cite{openmp3.1} is a widely used solution for Fortran, C
and C++ to describe loop-based and task-based parallelism. It consists of a few
directives attached to the code, directives that describe parallel loops and parallel code
sections in a shared memory model.

@@

The following listing gives a simple example of explicit loop-based parallelism
(OpenMP 3.0 task-based parallelism form is also supported, the interested
reader may refer to~\cite{pyhpc2013} for more details). the \texttt{kernel}
function is pure and computation intensive, making the loop a perfect candidate
for parallelization.

\begin{lstlisting}
def compute_julia(cr, ci, N, bound=1.5, lim=1000., cutoff=1e6):
    julia = np.empty((N, N), np.uint32)
    grid_x = np.linspace(-bound, bound, N)
    #omp parallel for 
    for i, x in enumerate(grid_x):
        for j, y in enumerate(grid_x):
            julia[i,j] = kernel(x, y, cr, ci, lim, cutoff)
    return julia
\end{lstlisting}

As in in standard OpenMP, loop indices of the parallel loop are automatically
made private, and Pythran extends the concept to any random access iterator,
even after type destructuring. In the example, both \texttt{i} and \texttt{x}
are private. A scope analysis is used to prove that \texttt{j} and \texttt{y}
are also local to the loop. In the end, the whole computation achieves a
perfect speedup of $\times4$

Next section performs an in-depth comparison of Pythran with others Python
optimizers: Cython and PyPy.

@@

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expression is still note mature for numpy expression, but there is a
strong on-going effort of the community to fix this.

@@

The ``blacksholes'' application is a benchmarks that showcases the
translation cost from the Python space to native space. Indeed we forced the
use of a \texttt{list} as input of the kernel instead of a numpy array. As a
consequence there is a costly translation step from a structure that contains
an array of pointers to generic Python objects to a dense structure involved.
Results are shown in Table~\ref{tbl:black}. They show that conversion code is
less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
it does not suffer from a similar conversion penalty.

@@

This paper gives an overview of the compilation flow, the analyses involved and
the optimizations used. It also compares the performance of compiled Pythran
modules against CPython and other optimizers like Cython and PyPy.

To conclude, limiting Python to a statically typed subset does not hinders the
expressiveness when it comes to scientific or mathematic computations, but makes
it possible to use a wide variety of classical optimizations to help Python
match the performances of statically compiled language. Moreover, one can use
high level information to generate efficient code that would be difficult to
write for the average programmer.

@@

This project has been partially funded by the CARP Project and the SILKAN
Company.  The authors would like to thank Yuancheng Peng and Eliott Coyac for
their contributions to development of the Pythran compiler.
